### 2020-11-11 latex/vlp.tex
Edits to Manfredi's improvements of VLP proof
Changing regular ref to eqref for equations

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The Cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian (\ref{eqn:simply-actuated-hamiltonian}) whose
dynamics (\ref{eqn:simply-actuated-dynamics}) resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting collinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

@@

The total mechanical energy of the system restricted to the constraint manifold
is given by (\ref{eqn:vlp-energy}), where \(L(q,p)\) and \(\dot{L}(q,p)\) are
known.
\begin{equation}\label{eqn:vlp-energy}
   E(q,p) = \frac{1}{2} \frac{p^2}{mL^2} + \frac{1}{2}\dot{L}^2 - mgL\cos(q)
   .
\end{equation}
In the rest of this chapter, we will derive a \(C^2\) function \(L(q,p)\) based
on natural human motion.
This function will produce constrained dynamics that inject energy into the VLP.

@@

Figure \ref{fig:vlp-optimal-controller-qp-plane} displays \(L^\star(q,p)\) 
on the \((q,p)\)-plane.
Note that the length remains constant inside each quadrant and changes only when it
crosses one of the axes. Using this fact, we can redefine
the time-optimal pumping strategy as a function of 
\(\theta := \arctan_2(p,q)\).
Abusing notation, we denote this by \(L^\star(\theta)\), which is defined in
(\ref{eqn:vlp-optimal-controller}).
Figure \ref{fig:vlp-optimal-controller} shows the graph of \(L^\star(\theta)\), 
where now the length varies between \([\underline{l},\overline{l}]\) rather than 
\(\{-1,0,1\}\).

@@

This family of constraints approximates \(L^\star(\theta)\) because 
\[
   \lim\limits_{T \rightarrow 0} L_T(\theta) = L^\star(\theta)
   .
\]
Unfortunately, while \(L_T(\theta)\) is continuously-differentiable, 
it is not twice-differentiable for most values of \(T\).
If we wish to use it as a VNHC, we must ensure that either the generalized
forces \(\tau\) acting on \(p_l\) can be discontinuous (which is certainly
achievable by humans), or we must find a value of \(T\) where this constraint is
at least \(C^2\).
Thankfully, setting \(T = \frac{\pi}{2}\) yields the smooth function 
\(L_{\frac{\pi}{2}}(\theta)\), which can be simplified from
(\ref{eqn:vlp-T-controller}) into
\begin{equation}\label{eqn:vlp-smoothed-controller}
   L_\frac{\pi}{2}(\theta) = -\Delta l \sin(2\theta) + l_{\text{avg}}
   .
\end{equation}
This smooth constraint is plotted for demonstration in Figure
\ref{fig:vlp-smoothed-controller}.

@@

Before proving the VLP will gain energy under this VNHC, 
we must rigorously define what that means.
If any initial condition of the constrained dynamics converges to the point
\((q,p) = (0,0)\), the momentum must be decreasing over time, so the VLP will
clearly be losing energy.
A notion of energy injection must therefore require that the origin repels all
solutions.
We need to avoid closed orbits or limit cycles in the \((q,p)\)-plane,
so that the system can always reach higher speeds. 
Finally, we want the VLP's momentum to increase on average, regardless of
its initial value.
This intuition is captured mathematically by the following definition.

\begin{defn}\label{defn:energy-gain}
   Let \(D \subset \R^2\) be open with \(0 \in D\).
   Let \(f : D \rightarrow \R^2\) be locally Lipschitz with \(f(0) = 0\).
   The system described by \(\dot{x} = f(x)\) 
   \textit{gains energy on \(D\)} if 
   \begin{enumerate}
      \item The origin is a repeller 
         (or equivalently, the origin is asymptotically stable in negative time).
      \item For all compact sets \(K \subset D\) and for almost every initial
         condition \(x(0) \in K\), there exists \(T > 0\) such
         that \(x(t) \not \in K \, (\forall t > T)\).
   \end{enumerate}
   The system \textit{loses energy on \(D\)} if it gains energy in
   negative-time.
\end{defn}

The second property of Definition \ref{defn:energy-gain} allows the system
to have unstable equilibria, but not limit cycles nor closed orbits.
The next definition ties this notion of energy gain to VNHCs.

\begin{defn}\label{defn:energy-injection}
   A regular VNHC \(h(\mathbf{q},\mathbf{p}) = 0\) \textit{injects (dissipates)
     energy} if the constrained dynamics in~\eqref{eqn:vlp-hamiltonian} gain
   (lose) energy everywhere on the constraint manifold, except possibly on a set
   of measure zero.
\end{defn}

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at $\theta \in
   \{0,\pi/2,\pi,3 \pi/2\}$.  If instead 
\[
 \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \leq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
\]
the VNHC dissipates energy.
\end{thm}
\begin{proof}
   We will first show that, under the given assumptions, the origin of the
   constrained dynamics~\eqref{eqn:vlp-hamiltonian} is a repeller, and for that
   we consider the negative time system obtained
   from~\eqref{eqn:vlp-hamiltonian} by reversing the sign of the vector field:
\begin{equation}\label{eqn:negative-time-vlp}
   \begin{cases}
      \dot{q} = -\frac{p}{m L^2} \\
      \dot{p} = mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
The state space of~\eqref{eqn:negative-time-vlp} is \(\mathbb{S}^1
\times \R\), and the system has two equilibria, \((q,p)=(0,0)\) and
\((q,p)=(\pi,0)\). We need to show that the equilibrium
\((q,p)=(0,0)\) is asymptotically stable
for~\eqref{eqn:negative-time-vlp}.

Consider the Lyapunov function candidate
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      ,
   \]
which corresponds to the energy of VLP when the pendulum length is fixed at
\(l_\text{avg}\).  The function \(E_\text{avg}\) is positive definite at
\((0,0)\), and has compact sublevel sets on \(\mathbb{S}^1 \times \R\).  The
derivative of \(E_\text{avg}\) along~\eqref{eqn:negative-time-vlp} is
   %% \begin{equation}\label{eqn:eavg-dot-pos}
   %%    \dot{E}_\text{avg}^+ = -\frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
   %%            {l_\text{avg}^2L(\theta)^2}
   %%    ,
   %% \end{equation}
   %% and the derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \begin{equation}\label{eqn:eavg-dot-neg}
      \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 -
        l_\text{avg}^3\right)} {l_\text{avg}^2L(\theta)^2}
      .
   \end{equation}
It is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       ,
   \]
and therefore,
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
          \nonumber \\
          &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
          \nonumber \\
          &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
          \nonumber \\
          &\leq 0 \text{ (by assumption)}
      .
   \end{align}
We have thus shown that \(\dot{E}_\text{avg} \leq 0\), and therefore
the equilibrium \((q,p)=(0,0)\) is stable for~\eqref{eqn:negative-time-vlp}. We
now apply the Krasovskii-LaSalle invariance principle, and consider the largest
subset of \( Z = \left\{(q,p) \in \mathbb{S}^1 \times \R \mid
\dot{E}_\text{avg}(q,p) = 0 \right\} \). We see from~\eqref{eqn:eavg-dot-neg}
that \(\dot{E}_\text{avg}(q,p)\) is zero when $p=0$, \(\sin(q)=0\), or
$L(\theta) = l_\text{avg}$. This latter condition by assumption is met when
\(\theta \in \{0,\pi/2,\pi, 3\pi/2\}\), or equivalently when either $q=0$ or
$p=0$. All in all, we have that
   \[
      Z= \left\{(q,p) \in \mathbb{S}^1 \times \R \mid q =0\right\} \cup \left\{(q,p) \in \mathbb{S}^1 \times \R \mid p =0\right\} 
      .
   \]
It is easily seen that the largest invariant subset of \(Z\) is the union of the two equilibria
\[
\Omega =  \{(0,0)\}\cup \{(\pi,0)\}
.
\]  
Since all sublevel sets of \(E_\text{avg}\) are compact, the
Krasovskii-LaSalle invariance principle implies that the set
\(\Omega\) is globally attractive. Since the two equilibria are
isolated, and since \((q,p)=(0,0)\) is stable, we deduce that the
equilibrium \((q,p)=(0,0)\) is asymptotically stable for the negative
time system~\eqref{eqn:negative-time-vlp}, and thus it is a repeller
for the constrained dynamics in~\eqref{eqn:vlp-hamiltonian}.

We now prove that almost all solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} escape compact sets in finite
time, i.e., that for each compact subset \(K\) of \(\mathbb{S}^1
\times \R\), and for almost every condition \( (q_0,p_0) \in K \),
there exists \( T > 0\) such that \( (q(t),p(t)) \not \in K \) for all
\( t > T \), where \( (q(t),p(t)) \) denotes the solution
of~\eqref{eqn:vlp-hamiltonian} with \( (q(0),p(0)) = (q_0,p_0)\). In
what follows, we denote by \( x = (q,p) \) the state of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}. Also, we denote by \(\Pi^-\) the
stable manifold of the equilibrium \((\pi,0)\) for
system~\eqref{eqn:vlp-hamiltonian}, defined as
 \[
      \Pi^- := \left\{ x(0) \in \mathbb{S}^1 \times \R \mid
      \lim_{t \to \infty} x(t) = (\pi, 0) \right\}.
   \]
The linearization of system~\eqref{eqn:vlp-hamiltonian} at the
equilibrium \((\pi,0)\) has a system matrix given by
  \begin{equation}\label{eqn:linearized-vlp}
      \begin{bmatrix}
         0 & -\frac{1}{mL(0)^2} \\
         -m g L(0) & 0 \\
      \end{bmatrix}
      ,
   \end{equation}
whose spectrum is \(\{\pm \sqrt{{g}/{L(0)}}\}\).  Since one eigenvalue
of (\ref{eqn:linearized-vlp}) is positive and one is negative, the
stable manifold theorem implies that the stable manifold \( \Pi^- \)
is a one-dimensional immersed submanifold of \(\mathbb{S}^1 \times
\R\), and thus a set of zero measure.

Let \( K \subset \mathbb{S}^1 \times \R\) be an arbitrary compact set
and let \( x(0) \in K \backslash \left(\Pi^- \cup
\left\{(0,0)\right\}\right)\). Since \( x(0) \not \in \Pi^- \), we have that
\begin{equation}\label{eqn:vlp-x-does-not-converge}
x(t) \not \to_{t \to \infty} (\pi,0).
\end{equation}
Suppose, by way of contradiction, that for each \(T
> 0\) there exists \(t' > T\) such that \(x(t') \in K\).  Letting
   \[
      k := \max\limits_{x \in K} E_\text{avg}(x)
      ,
   \] 
and \(E_k =\{ x \mid E_\text{avg}(x) \leq k\} \), we have that \( K
\subset E_k \), and thus \(x(t') \in E_k\). We have shown earlier that
the function \( E_\text{avg} \) is nonincreasing along solutions
of~\eqref{eqn:negative-time-vlp}, which implies that \(
E_\text{avg}(x(t)) \) is nondecreasing for solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}, and therefore the half orbit
\(x([0,t'])\) is contained fully within \(E_k\).  In particular,
\(x(T) \in E_k\). Since this is true for each \( T >0 \), we deduce
that \(x(t) \in E_k\) for all \(t \in [0,\infty[\).  Since \(E_k\) is
    compact, \(x(t)\) has a positive limit set in \(E_k\) which, by
    the Poincar\'{e}-Bendixson theorem, is either an equilibrium or a
    closed orbit \cite{poincare_bendixson}.  We will show this
    positive limit set must be an equilibrium and thereby reach a
    contradiction.

In order to rule out closed orbits, suppose that the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} have a periodic solution \(
z(t)\). Since \( E_\text{avg}(z(t)) \) is nondecreasing and \(z(t)\)
is periodic, it must be that \( E_\text{avg}(z(t)) \) is constant, or
\(\dot E_\text{avg}(z(t)) \equiv 0\). We have shown earlier
that\footnote{We have shown it for the negative time
  system~\eqref{eqn:negative-time-vlp}, but changing the sign of the
  vector field does not change the set where \(\dot E_\text{avg}
  =0\) nor the invariant sets.}
\[
\{x \mid \dot E_\text{avg}(x)=0\} = \left\{(q,p) \mid q =0\right\}
\cup \left\{(q,p) \mid p =0\right\},
\]
and the only invariant subset of this set is the union of two disjoint
equilibria. Since the orbit \(z(\R)\) is an invariant set which is
connected, \(z(\R)\) must be an equilibrium and cannot be a nontrivial
closed orbit.

@@

We conclude that the VNHC injects energy into the VLP.
   By flipping the inequality of (\ref{eqn:vlp-energy-gain-condition})
   we find the VLP is gaining energy in negative-time, so the VNHC is
   dissipating energy.
\end{proof}

>>>>

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The Cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian ~\eqref{eqn:simply-actuated-hamiltonian} whose
dynamics ~\eqref{eqn:simply-actuated-dynamics} resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting collinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

@@

The total mechanical energy of the system restricted to the constraint manifold
is given by ~\eqref{eqn:vlp-energy}, where \(L(q,p)\) and \(\dot{L}(q,p)\) are
known.
\begin{equation}\label{eqn:vlp-energy}
   E(q,p) = \frac{1}{2} \frac{p^2}{mL^2} + \frac{1}{2}\dot{L}^2 - mgL\cos(q)
   .
\end{equation}
In the rest of this chapter, we will derive a \(C^2\) function \(L(q,p)\) based
on natural human motion.
This function will produce constrained dynamics that inject energy into the VLP.

@@

Figure \ref{fig:vlp-optimal-controller-qp-plane} displays \(L^\star(q,p)\) 
on the \((q,p)\)-plane.
Note that the length remains constant inside each quadrant and changes only when it
crosses one of the axes. Using this fact, we can redefine
the time-optimal pumping strategy as a function of 
\(\theta := \arctan_2(p,q)\).
Abusing notation, we denote this by \(L^\star(\theta)\), which is defined in
~\eqref{eqn:vlp-optimal-controller}.
Figure \ref{fig:vlp-optimal-controller} shows the graph of \(L^\star(\theta)\), 
where now the length varies between \([\underline{l},\overline{l}]\) rather than 
\(\{-1,0,1\}\).

@@

This family of constraints approximates \(L^\star(\theta)\) because 
\[
   \lim\limits_{T \rightarrow 0} L_T(\theta) = L^\star(\theta)
   .
\]
Unfortunately, while \(L_T(\theta)\) is continuously-differentiable, 
it is not twice-differentiable for most values of \(T\).
If we wish to use it as a VNHC, we must ensure that either the generalized
forces \(\tau\) acting on \(p_l\) can be discontinuous (which is certainly
achievable by humans), or we must find a value of \(T\) where this constraint is
at least \(C^2\).
Thankfully, setting \(T = \frac{\pi}{2}\) yields the smooth function 
\(L_{\frac{\pi}{2}}(\theta)\), which can be simplified from
~\eqref{eqn:vlp-T-controller} into
\begin{equation}\label{eqn:vlp-smoothed-controller}
   L_\frac{\pi}{2}(\theta) = -\Delta l \sin(2\theta) + l_{\text{avg}}
   .
\end{equation}
This smooth constraint is plotted for demonstration in Figure
\ref{fig:vlp-smoothed-controller}.

@@

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at 
   \(\theta \in \{0, \frac{\pi}{2}, \frac{\pi}, \frac{3\pi}{2}\}\).  
   If instead 
   \[
    \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \leq 0 
         \text{ }\forall \theta \in \mathbb{S}^1
         ,
   \]
   the VNHC dissipates energy.
\end{thm}
\begin{proof}
We will first show that, under the given assumptions, the origin of the
constrained dynamics~\eqref{eqn:vlp-hamiltonian} is a repeller, and for that
we consider the negative time system obtained
from~\eqref{eqn:vlp-hamiltonian} by reversing the sign of the vector field:
\begin{equation}\label{eqn:negative-time-vlp}
   \begin{cases}
      \dot{q} = -\frac{p}{m L^2} \\
      \dot{p} = mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
The state space of~\eqref{eqn:negative-time-vlp} is 
\(\mathbb{S}^1 \times \R\), and the system has two equilibria,
\((q,p) = (0,0)\) and \((q,p) = (\pi,0)\). 
We need to show that the equilibrium \((q,p) = (0,0)\) is
asymptotically stable for~\eqref{eqn:negative-time-vlp}.

Consider the Lyapunov function candidate
\[
    E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
        + m g l_\text{avg} (1-\cos(q))
   ,
\]
which corresponds to the energy of VLP when the pendulum length is fixed at
\(l_\text{avg}\). 
The function \(E_\text{avg}\) is positive definite at \((0,0)\), and has compact
sublevel sets on \(\mathbb{S}^1 \times \R\).  
The derivative of \(E_\text{avg}\) along~\eqref{eqn:negative-time-vlp} is
\begin{equation}\label{eqn:eavg-dot-neg}
   \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 -
     l_\text{avg}^3\right)} {l_\text{avg}^2L(\theta)^2}
   .
\end{equation}
It is easy to show that
\[
   \sign{\sin(q)p} = \sign{\sin(2\theta)}
   .
\]
Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
\[ 
    \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
    ,
\]
and therefore,
\begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= 
       \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
    \nonumber \\
    &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
    \nonumber \\
    &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
    \nonumber \\
    &\leq 0 \text{ (by assumption)}
   .
\end{align}
We have thus shown that \(\dot{E}_\text{avg} \leq 0\), and therefore
the equilibrium \((q,p)=(0,0)\) is stable for~\eqref{eqn:negative-time-vlp}.
We now apply the Krasovskii-LaSalle invariance principle, and consider the
largest subset of 
\( Z = \left\{(q,p) \in \mathbb{S}^1 \times \R \mid \dot{E}_\text{avg}(q,p) = 0
\right\} \). 
We see from~\eqref{eqn:eavg-dot-neg}
that \(\dot{E}_\text{avg}(q,p)\) is zero when \(p = 0\), \(\sin(q) = 0\), or
\(L(\theta) = l_\text{avg}\). 
This latter condition by assumption is met when 
\(\theta \in \{0,\pi/2,\pi, 3\pi/2\}\), or equivalently when either 
\(q = 0\) or \(p = 0\). 
All in all, we have that
\[
   Z= \left\{(q,p) \in \mathbb{S}^1 \times \R \mid q =0\right\} 
   \cup \left\{(q,p) \in \mathbb{S}^1 \times \R \mid p =0\right\} 
   .
\]
It is easily seen that the largest invariant subset of \(Z\) is the union of the
two equilibria
\[
   \Omega =  \{(0,0)\}\cup \{(\pi,0)\}
   .
\]  
Since all sublevel sets of \(E_\text{avg}\) are compact, the
Krasovskii-LaSalle invariance principle implies that the set
\(\Omega\) is globally attractive \cite{krasovskii_lasalle}. 
Since the two equilibria are isolated, and since \((q,p)=(0,0)\) is stable, we
deduce that the equilibrium \((q,p)=(0,0)\) is asymptotically stable for the
negative time system~\eqref{eqn:negative-time-vlp}, and thus it is a repeller
for the constrained dynamics in~\eqref{eqn:vlp-hamiltonian}.

We now prove that almost all solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} escape compact sets in finite
time, i.e., that for each compact subset \(K\) of 
\(\mathbb{S}^1 \times \R\), and for almost every condition 
\( (q_0,p_0) \in K \), there exists \( T > 0\) such that
\( (q(t),p(t)) \notin K \) for all \( t > T \), where \( (q(t),p(t)) \)
denotes the solution of~\eqref{eqn:vlp-hamiltonian} with 
\( (q(0),p(0)) = (q_0,p_0)\). 
In what follows, we denote by \( x = (q,p) \) the state of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}. 
Also, we denote by \(\Pi^-\) the stable manifold of the equilibrium \((\pi,0)\)
for system~\eqref{eqn:vlp-hamiltonian}, defined as
\[
   \Pi^- := \left\{ x(0) \in \mathbb{S}^1 \times \R \mid
   \lim_{t \to \infty} x(t) = (\pi, 0) \right\}.
\]
The linearization of system~\eqref{eqn:vlp-hamiltonian} at the
equilibrium \((\pi,0)\) has a system matrix given by
\begin{equation}\label{eqn:linearized-vlp}
   \begin{bmatrix}
      0 & -\frac{1}{mL(0)^2} \\
      -m g L(0) & 0 \\
   \end{bmatrix}
   ,
\end{equation}
whose spectrum is \(\{\pm \sqrt{{g}/{L(0)}}\}\).  
Since one eigenvalue of ~\eqref{eqn:linearized-vlp} is positive and one is
negative, the stable manifold theorem implies that the stable manifold 
\( \Pi^- \) is a one-dimensional immersed submanifold of
\(\mathbb{S}^1 \times \R\), and thus is a set of measure zero
\cite{stable_manifold}.

Let \( K \subset \mathbb{S}^1 \times \R\) be an arbitrary compact set and let 
\( x(0) \in K \backslash \left(\Pi^- \cup \left\{(0,0)\right\}\right)\).
Since \( x(0) \notin \Pi^- \), we have that
\begin{equation}\label{eqn:vlp-x-does-not-converge}
   x(t) \nrightarrow_{t \to \infty} (\pi,0).
\end{equation}
Suppose, by way of contradiction, that for each \(T > 0\) there exists 
\(t' > T\) such that \(x(t') \in K\).  
Letting
\[
   k := \max\limits_{x \in K} E_\text{avg}(x)
   ,
\] 
and \(E_k =\{ x \mid E_\text{avg}(x) \leq k\} \), we have that 
\( K \subset E_k \), and thus \(x(t') \in E_k\).
We have shown earlier that the function \( E_\text{avg} \) is nonincreasing
along solutions of~\eqref{eqn:negative-time-vlp}, which implies that
\( E_\text{avg}(x(t)) \) is nondecreasing for solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}, and therefore the half orbit \(x([0,t'])\)
is contained fully within \(E_k\).  
In particular, \(x(T) \in E_k\). Since this is true for each \( T >0 \), we
deduce that \(x(t) \in E_k\) for all \(t \in [0,\infty[\).  
Since \(E_k\) is compact, \(x(t)\) has a positive limit set in \(E_k\) which, by
the Poincar\'{e}-Bendixson theorem, is either an equilibrium or a closed orbit
\cite{poincare_bendixson}.  
We will show this positive limit set must be an equilibrium and thereby reach a
contradiction.

In order to rule out closed orbits, suppose that the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} have a periodic solution \( z(t)\). 
Since \( E_\text{avg}(z(t)) \) is nondecreasing and \(z(t)\)
is periodic, it must be that \( E_\text{avg}(z(t)) \) is constant, or
\(\dot E_\text{avg}(z(t)) \equiv 0\). 
We have shown earlier
that\footnote{We have shown it for the negative time
system~\eqref{eqn:negative-time-vlp}, but changing the sign of the
vector field does not change the set where 
\(\dot E_\text{avg} =0\) nor the invariant sets.}
\[
   \{x \mid \dot E_\text{avg}(x) = 0\} = \left\{(q,p) \mid q = 0\right\}
   \cup \left\{(q,p) \mid p =0\right\},
\]
and the only invariant subset of this set is the union of two disjoint
equilibria. 
Since the orbit \(z(\R)\) is an invariant set which is
connected, \(z(\R)\) must be an equilibrium and cannot be a nontrivial
closed orbit.

@@

We conclude that the VNHC injects energy into the VLP.
By flipping the inequality of ~\eqref{eqn:vlp-energy-gain-condition}
we find the VLP is gaining energy in negative-time, so the VNHC is
dissipating energy.
\end{proof}
